---
title: "Week 2: Starting with R"
author: "UZH n UU: tested by Jonas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=FALSE}
# install readr from CRAN
# install.packages("readr")
library(readr) # make sure your R has this library loaded, if not install it (see line above)
```

# Exercises for the R-Beginner - The fancy calculator

## Exercise 3
* from https://alexd106.github.io/intro2R/exercise_3.html

### Task 4

#### Observations (rows) → Each row is an individual station where data was collected.
#### Variables (columns) → Includes whale count, time spent at station, latitude, longitude, depth, seabed gradient, noise level, etc.
#### No missing data → Every row is complete.
#### Tidy format → Each column is a variable, and each row is an observation.


### Task 5

```{r}
# Import the mywhaledata.tsv file

getwd()
whale <- read.table("../data4exercises/mywhaledata.tsv", header = T, stringsAsFactors = T)

# Check the first few rows to confirm import
head(whale)

# Get a summary of the dataset
summary(whale)

# Check the structure of the dataset
str(whale)

# Check for missing values (should be 0)
sum(is.na(whale))

```

### Task 7

```{r}

# Get a summary of the dataset
summary(whale)

# Count missing values in each column
colSums(is.na(whale))

# One value is missing in "number of whales"

```

### Task 9

```{r}

# Extract rows where depth is greater than 1200 meters
depths_greater_than_1200 <- whale[whale$depth > 1200, ]
depths_greater_than_1200

# Extract rows where the gradient is steeper than 200 degrees
steeper_than_200 <- whale[whale$gradient > 200, ]
steeper_than_200

# Extract rows where the water noise level is 'low'
low_noise <- whale[whale$water.noise == "low", ]
low_noise

# Extract rows where the water noise level is 'high' and the month is 'May'
high_noise_may <- whale[whale$water.noise == "high" & whale$month == "May", ]
high_noise_may

# Extract rows where the month is 'October', water noise level is 'low', and gradient is greater than the median value of 132
october_low_noise_steep_gradient <- whale[whale$month == "October" & whale$water.noise == "low" & whale$gradient > 132, ]
october_low_noise_steep_gradient

# Extract rows where the latitude is between 60.0 and 61.0 and longitude is between -6.0 and -4.0
lat_lon_range <- whale[whale$latitude >= 60.0 & whale$latitude <= 61.0 & whale$longitude >= -6.0 & whale$longitude <= -4.0, ]
lat_lon_range

# Extract rows where the water noise level is not 'medium'
not_medium_noise <- whale[whale$water.noise != "medium", ]
not_medium_noise

```

### Task 10

```{r}

# Calculate the median of the gradient variable
median_gradient <- median(whale$gradient, na.rm = TRUE)
median_gradient

# Extract rows where the month is 'October', water noise level is 'low', and gradient is greater than the median value of gradient
october_low_noise_steep_gradient <- whale[whale$month == "October" & whale$water.noise == "low" & whale$gradient > median_gradient, ]
october_low_noise_steep_gradient

## OR

# Calculate the median of the gradient variable (R will handle NAs in the column by returning NA)
median_gradient <- median(whale$gradient)
median_gradient

# Extract rows where the month is 'October', water noise level is 'low', and gradient is greater than the median value of gradient
october_low_noise_steep_gradient <- whale[whale$month == "October" & whale$water.noise == "low" & whale$gradient > median_gradient, ]
october_low_noise_steep_gradient

```


### Task 11

```{r}
# Calculate the mean number of whales spotted, ignoring NA values
mean_whales <- mean(whale$number.whales, na.rm = TRUE)
mean_whales

# Extract rows where depth is greater than 1500 meters and number of whales spotted is greater than the calculated mean
depth_greater_1500_and_whales_greater_than_avg <- whale[whale$depth > 1500 & whale$number.whales > mean_whales, ]


head(depth_greater_1500_and_whales_greater_than_avg)

```
