---
title: "SwitchingGears_RomaneLamoity"
output: html_document
---

```{r}

knitr::opts_chunk$set(echo = TRUE)
install.packages("tidyverse", repos = "https://cloud.r-project.org/")
```

## Load Required Libraries
```{r load-libraries}
library(tidyverse) 
```

## Import Data
```{r import-data}
squid <- read.table("../data4exercises/squid1.txt", header = TRUE, stringsAsFactors = FALSE)
```

## Data Wrangling using Pipes
Convert necessary variables into factors and filter out missing values.
```{r data-wrangling}
squid <- squid %>%
  mutate(
    year_factor = as.factor(year),
    month_factor = as.factor(month),
    maturity_stage_factor = as.factor(maturity.stage)
  ) %>%
  drop_na()
```

## DML vs Weight Scatter Plot
Let's create a scatter plot showing the relationship between `DML` and `weight`.
```{r ggplot-scatter}
ggplot(squid, aes(x = DML, y = weight)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Scatter Plot of DML vs. Weight",
    x = "Dorsal Mantle Length (DML)",
    y = "Weight (g)"
  ) +
  theme_minimal()
```

## DML vs Weight Box Plot
```{r boxplot-maturity}

ggplot(squid, aes(x = factor(maturity.stage), y = DML, fill = factor(maturity.stage))) +
  geom_boxplot() +
  labs(
    title = "DML Distribution Across Maturity Stages",
    x = "Maturity Stage",
    y = "Dorsal Mantle Length (DML)"
  ) +
  theme_minimal()
```

## Occupational Group vs Smoking Mortality Bar Chat
```{r barchart-smoking}
getwd()
smoking <- read.table("../data4exercises/smoking.txt", header = TRUE, stringsAsFactors = TRUE) 

smoking_summary <- smoking%>%
    group_by(occupational.group) %>%
    summarise(avg_smoking = mean(smoking, na.rm = TRUE))

ggplot(smoking_summary, aes(x = occupational.group, y = avg_smoking, fill = occupational.group)) + 
    geom_bar(stat = "identity") + 
    labs(title = "Occupational Group vs Smoking Mortality", 
         x = "Occupational Group", 
         y = "Smoking Mortality"
         ) + 
    theme_minimal()
```

## Occupational Group vs Smoking Mortality Boxplot
```{r barchart-smoking-boxplot}
getwd()
smoking <- read.table("../data4exercises/smoking.txt", header = TRUE, stringsAsFactors = TRUE) 
ggplot(smoking, aes(x = occupational.group, y = smoking, fill = occupational.group)) +
    geom_boxplot() +
    labs(title = "Occupational Group vs Smoking Mortality", 
         x = "Occupational_Group",
         y = "Smoking Mortality"
    ) + 
    theme_minimal()
```

## Speed vs Distance
```{r barchart-cars}
getwd()
smoking <- read.table("../data4exercises/cars.txt", header = TRUE, stringsAsFactors = TRUE) 
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point(color = "blue") +  
  geom_smooth(method = "lm", se = FALSE, color = "red") + 
  labs(title = "Speed vs Distance",
       x = "Speed (mph)",
       y = "Stopping Distance (ft)") +
  theme_minimal()
```
##Pollution
```{r boxplot-atmosphere}

atmosphere <- read.table("../data4exercises/atmosphere.txt", header = TRUE, stringsAsFactors = FALSE)

ggplot(atmosphere, aes(x = factor(treatment), y = moisture, fill = factor(treatment))) +
  geom_boxplot() +
  labs(
    title = "Moisture in different Treatments",
    x = "TREATMENT",
    y = "MOISTURE"
  ) +
  theme_minimal()
```
